Supp mat: time series paper

### 2017-07-07 writing/supp_mat_draft.tex
cleaned up supp mat

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{float} \renewcommand{\thepage}{S\arabic{page}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

@@

Code for all the figures and tables can be found at
\href{https://github.com/erwhite1}{}.

@@

Here, we illustrate in detail how we performed the subsampling and power
calculations for a specific population. As an example, we examine a
35-year time series of (). Simple linear regression indicates a
significant decline for this population with a blank and blank. We
assume that this significant decline over 35 years is in fact the ``true
trend``. In statistical jargon, the 35-year trend is an effect that is
actually present. We can then use this as a benchmark to see if
subsamples of the time series show a similar result.

We first extract all contiguous subsamples of the time series. This
leads to 34 two-year subsamples, 33 two-year subsamples, and so forth
until a single 35-year subsample. We can call each set of subsamples, of
the same length, a set. For each subsample, we conduct linear regression
and extract model coefficients and p-values. Then, the fraction of
subsamples within a set that show significant trends (significant slope
coefficient) is the statistical power. It is important to note that we
only consider subsamples to be significant if they are significant in
the same direction as the complete 35-year time series.

We can then plot statistical power as a function of time series length.
As expected, we can see that power increases with the more years that
are sampled.

Then, we determine an appropriate level of statistical power that we
find acceptable. Traditionally, this has been at 0.8, however, this is
purely historical. Statistical power of 0.8 implies\ldots{}

With statistical power of 0.8, we then determine the minimum time series
length (\(t_min\)) required to achieve that level of statistical power.
Here, \(t_min\) is the first point as which all points to the right are
above 0.8.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{(a) Population size of Harp Seals over time. The line is the
best fit line from linear regresssion. (b) Statistical power for
different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

In Figure \ref{fig:empirical_approach_example}b, statistical power
increases with more time sampled until our desired power of 0.8 at 16
years. Therefore, 16 years would be our estimate for \(T_{min}\).

@@

In the main text, we explained how the minimum time required strongly
correlated with the trend strength, temporal autocorrelation, and
variance in population size. Here we use a generalized linear model
framework with a Poisson error structure to determine drivers of the
minimum time required. In figure, \ref{fig:poisson_model} we show a set
of residual plots for the regression. We then show the coefficient
estimates and levels of significance in table \ref{table:model_output}.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/poisson_model-1.pdf}
\caption{Output of poisson regression\ldots{}\label{fig:poisson_model}}
\end{figure}

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Stuff here\label{table:model_output}}\tabularnewline
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endfirsthead
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endhead
(Intercept) & 3.7141297 & 0.0274350 & 135.379203 &
0.0000000\tabularnewline
abs\_overall\_trend & -104.4974288 & 3.0997732 & -33.711314 &
0.0000000\tabularnewline
autocorrelation & -0.4905888 & 0.0431062 & -11.380936 &
0.0000000\tabularnewline
variance & 18.7756472 & 0.6409982 & 29.291262 & 0.0000000\tabularnewline
gen\_len & 0.0053584 & 0.0025995 & 2.061352 & 0.0392695\tabularnewline
\bottomrule
\end{longtable}

Our model with trend strength, autocorrelation, variance, and generation
length accounted for 72.55\% of the variation in the minimum time
required (Table \ref{table:model_output}). However, we also found trend
strength and variance to be strongly correlated with one another.
Therefore, we ran two additional models with either trend strength or
variance, but not both together. This resulted in lower explained
deviance (analogue to \(R^2\)) of 53.08\% and 45.04\%, respectively.

@@

In the main manuscript, we examined the minimum time required to detect
a significant trend in abundance over time using linear regression. As
detailed in the main manuscript the minimum time required was around 15,
but there was a wide distribution. Therefore, we were interested in
potential explanatory variables of the minimum time required. In the
main manuscript, we examined characteristics of the time series itself,
like variability, autocorrelation, and the trend in abundance over time.
Here, we combined our time series database with a database on life
history characteristics of amniotes from (cite other database here).
There was life history information available for 547 populations
representing 315 different species, all of which were in the Aves class.

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{(a) Minimum time required to estimate change in abundance for
species class, (b) longterm trend (estimated slope coefficient) by
species class, (c) interannual variability in population size by species
class, and (d) temporal autocorrelation by species
class.\label{fig:class}}
\end{figure}

@@

\subsection{\texorpdfstring{Sensitivity analysis of \(\alpha\) and
power}{Sensitivity analysis of \textbackslash{}alpha and power}}\label{sensitivity-analysis-of-alpha-and-power}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{Minimum time required to assess longterm trends in abundance
for values of statistical significance (\(\alpha\)) and power
(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
\end{figure}

@@

\section{Results from other questions of
interest}\label{results-from-other-questions-of-interest}

\subsection{Simulations with more complicated population
model}\label{simulations-with-more-complicated-population-model}

In the main text, we showed how a simple population model could be
simulated repeatedly to estimate the power obtained with time series of
increasing length. The model in the main text simulated linear
population growth with only a slope coefficient, y-intercept, and noise
parameter required. This model is purely phenomenological and does not
include any species life history. Here, we use the same routine as the
main text, but simulate from a more biologically-realistic population
model, the (MODEL NAME HERE):

\begin{equation}
    y = x
  \end{equation}

We simulate this model repetitively for different lengths of time
series. Statistical power is the fraction of time series of a set length
that showed significant trends in abundance, given a true trend in
abundance. \emph{How do I determine significance here? Through linear
regression again, or do I look at non-zero estimation of the parameter
values?}

\subsection{Empirical approach with more complicated population model
(use model similar to de Valpine
work?)}\label{empirical-approach-with-more-complicated-population-model-use-model-similar-to-de-valpine-work}

\subsection{Growth rate calculations}\label{growth-rate-calculations}

Instead of detecting a trend over time with linear regression, we could
also calculate the geometric growth rate of the population. In figure
\ref{fig:growth_rate}, we show how to calculate growth rates for
subsamples of a time series. First, we created subsamples of each
possible length from the full 35 year time series, as we did in the main
next. Next, we calculated the mean and standard deviation of growth
rates for each possible time series length (Fig.
\ref{fig:growth_rate}b). Lastly, we calculated the percent error
\(\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|\)
between the mean of each time series length (observed) and the overall
population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c),
we show the percent error as a function of time series length. Here we
define the minimum time required as the minimum number of years to
achieve less than 10\% error.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Example of calculating minimum time required for growth rate
estimation. (a) abundance over time, (b) mean and standard deviation of
growth rate for subsamples of entire time series, and (c) the percent
error between mean estimated growth rate and the true longterm growth
rate. The vertical bar denotes the minimum time required to estimate
growth rate within 10\% error.\label{fig:growth_rate}}
\end{figure}

We applied the same calculations to 1032 populations, as we did in the
main text. We then obtain a distribution of the minimum time required to
measure the ``true" longterm growth rate (Fig.
\ref{fig:min_time_growth_dist}). We see a bimodal distribution with many
populations required 30+ years to estimate the longterm growth rate. The
large number of short year required is due to cases where the entire
time series is consistently increasing or decreasing at the same rate.

@@

\subsection{Using Generalized additive model to identify significant
trends}\label{using-generalized-additive-model-to-identify-significant-trends}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{results here from GAM.\label{fig:gam_example}}
\end{figure}

We then fit GAM models for XX populations. We found\ldots{} (Fig.
\ref{fig:gam_result}). This is shorter/longer than the main text
question.

\subsection{Minimum time to detect density-dependence in time
series}\label{minimum-time-to-detect-density-dependence-in-time-series}

For decades, the importance of density-dependence in natural populations
has been an important question in ecology (citations). To address this
question, \citet{Knape2012} used data from the Global Population
Dynamics Database \citep{GPDD2010}. They fitted 627 time series with a
Gompertz population model and found evidence for density-dependence in
45\% of the time series. Each time series in their database contained at
least 15 data points, and ranged from 15 to 157. A follow up question to
their work would be, what time series length is required to confidently
estimate density-dependence.

@@

\renewcommand\refname{References}
\bibliography{White_bib.bib}

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{float} \renewcommand{\thepage}{S\arabic{page}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}} \usepackage[round]{natbib}
\bibpunct[; ]{(}{)}{,}{a}{}{;}

@@

\vspace{1cm}

Code for all the figures and tables can be found at
(\url{https://github.com/erwhite1}).

@@

Here, we illustrate how we performed the subsampling and power
calculations for a specific population. We provide more detail here than
in the main manuscript. As an example, we examine a 35-year time series
of Bigeye tuna (\emph{Thunnus obesus}), one species in the Global
Population Dynamics Database (NERC Centre for Population Biology 2010).
We assume that this significant increase over 35 years is in fact the
``true trend``. In statistical jargon, the 35-year trend is an effect
that is actually present, we can reject the null hypothesis of no trend.
We can then use this as a benchmark to see if subsamples of the time
series also indicate a significant increase.

We first extract all contiguous subsamples of the time series. This
leads to 34 two-year subsamples, 33 two-year subsamples, and so forth
until a single 35-year subsample. We can call each set of subsamples, of
the same length, a set. For each subsample, we conduct linear regression
and extract model coefficients and p-values. Then, the fraction of
subsamples within a set that show significant trends (significant slope
coefficient) is the statistical power. It is important to note that we
only consider subsamples to be significant if they are significant in
the same direction as the complete 35-year time series. In other words,
we are conducting a one-tailed test.

We can then plot statistical power as a function of time series length
(Fig. \ref{fig:empirical_approach_example}b). As expected, we can see
that power increases with the more years that are sampled. Then, we
determine an appropriate level of statistical power that we find
acceptable. Traditionally, this has been at 0.8, however, this is purely
historical. Statistical power of 0.8 implies if a true trend is present,
or a real change in abundance, then we will detect this trend 0.8
fraction of the time.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
time. The line is the best fit line from linear regresssion. (b)
Statistical power for different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

With statistical power of 0.8, we then determine the minimum time series
length (\(T_{min}\)) required to achieve that level of statistical
power. Here, \(T_{min}\) is the first point in Fig.
\ref{fig:empirical_approach_example}b where following points are also
above 0.8. In this example, \(T_{min}\) is 22. Therefore, a minimum of
22 years of continuous monitoring are required to have 0.8 statistical
power in determining long-term changes in abundance.

@@

In the main text, we explained how the minimum time required strongly
correlated with the trend strength, temporal autocorrelation, and
variance in population size. Here, we use a generalized linear model
framework with a Poisson error structure to determine drivers of the
minimum time required. In figure, \ref{fig:poisson_model} we show a set
of residual plots for the regression. We then show the coefficient
estimates and levels of significance in table \ref{table:model_output}.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/poisson_model-1.pdf}
\caption{Output of generalized linear model with a Poisson error
structure for predicting the minimum time required with explanatory
variables of the absolute value of the slope coefficient (or trend
strength), temporal autocorrelation, and variability in population
size.\label{fig:poisson_model}}
\end{figure}

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Output of generalized linear model to examine predictors of the
minimum time required for determing long-term population
trends.\label{table:model_output}}\tabularnewline
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endfirsthead
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endhead
(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
0\tabularnewline
autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
\bottomrule
\end{longtable}

Our model with trend strength, autocorrelation, variance, and generation
length accounted for 72.45\% of the variation in the minimum time
required (Table \ref{table:model_output}). However, we also found trend
strength and variance to be strongly correlated with one another.
Therefore, we ran two additional models with either trend strength or
variance, but not both together. This resulted in lower explained
deviance (analogue to \(R^2\)) of 53.03\% and 45.03\%, respectively.

@@

In the main manuscript, we examined the minimum time required to detect
a significant trend in abundance over time using linear regression. As
detailed in the main manuscript the minimum time required was around 15,
but there was a wide distribution. Therefore, we were interested in
potential explanatory variables of the minimum time required. In the
main manuscript, we examined characteristics of the time series itself,
like variability, autocorrelation, and the trend in abundance over time.
Here, we combined our time series database with a database on life
history characteristics of amniotes from Myhrvold et al. (2015). There
was life history information available for 547 populations representing
315 different species, all of which were in the Aves class.

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{(a) Minimum time required to estimate change in abundance for
species class, (b) long-term trend (estimated slope coefficient) by
species class, (c) interannual variability in population size by species
class, and (d) temporal autocorrelation by species
class.\label{fig:class}}
\end{figure}

@@

\subsection{Sensitivity analysis of significance level and
power}\label{sensitivity-analysis-of-significance-level-and-power}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{Minimum time required to assess long-term trends in abundance
for values of statistical significance (\(\alpha\)) and power
(\(1-\beta\)).\label{fig:min_time_vs_alpha_beta}}
\end{figure}

\section{Minimum time calculations testing exponential
growth}\label{minimum-time-calculations-testing-exponential-growth}

In the main text we evaluated the minimum time required to determine
long-term trends in abundance via linear regression. This process
examined linear trends in abundance over time. Here we examine the
minimum time required to estimate long-term trends that are either
exponential growth or decay. We use the same methods as described in the
previous sections, but we take the \(log\) of population density, or
abundance.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in log(abundance) given power of
0.8.\label{fig:min_time_dist_log_pop}}
\end{figure}

We see that the distribution of \(T_{min}\) is almost identical to that
in the main manuscript. This is perhaps not surprising as most time
series that would signficantly increase or decrease linearly would
probably also significantly increase or decrease at an exponential rate.
Further, the calculations here and in the main manuscript both use
linear regression, and both estimate the same number of parameters.

@@

\section{Simulations with more complicated population
model}\label{simulations-with-more-complicated-population-model}

In the main text, we showed how a simple population model could be
simulated repeatedly to estimate the power obtained with time series of
increasing length. The model in the main text simulated linear
population growth with only a slope coefficient, y-intercept, and noise
parameter required. This model is purely phenomenological and does not
include any species life history. Here, we use the same routine as the
main text, but simulate from a more biologically-realistic population
model. We simulate the model described in White, Nagy, and Gruber
(2014). The model is a stochastic, age-structured population model that
includes density-dependence for lemon sharks in Bimini, Bahamas.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Statistical power for different length of time series
simulations for a lemon shark population in Bimini,
Bahamas.\label{fig:shark_example}}
\end{figure}

Here, we parameterize the model for a situation where adult mortality
rate is high enough to cause a population decline. In the same way we
simulated models in the main text, we simulate this more
biologically-realistic model for different lengths of time. For each
length of time, we calculate the statistical power. Similar to results
with the simpler model, statistical power generally increases with
longer sampling time (Fig. \ref{fig:shark_example}). In this example,
the minimum time required (\(T_{min}\)) to obtain at least 0.8
statistical power, given a significance level of 0.05, is 27 years.

\section{Minimum time required to estimate geometric growth
rate}\label{minimum-time-required-to-estimate-geometric-growth-rate}

Instead of detecting a trend over time with linear regression, we could
also calculate the geometric growth rate of the population. In figure
\ref{fig:growth_rate}, we show how to calculate growth rates for
subsamples of a time series. First, we created subsamples of each
possible length from the full 35 year time series, as we did in the main
next. Next, we calculated the mean and standard deviation of growth
rates for each possible time series length (Fig.
\ref{fig:growth_rate}b). Lastly, we calculated the percent error
\(\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|\)
between the mean of each time series length (observed) and the overall
population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c),
we show the percent error as a function of time series length. Here we
define the minimum time required as the minimum number of years to
achieve less than 20\% error.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{Example of calculating minimum time required for growth rate
estimation. (a) European herring gull (\emph{Larus argentatus}) scaled
density over time, (b) mean and standard deviation of growth rate for
subsamples of entire time series, and (c) the percent error between mean
estimated growth rate and the true long-term growth rate. The vertical
bar denotes the minimum time required to estimate growth rate within
20\% error.\label{fig:growth_rate}}
\end{figure}

We applied the same calculations to 1032 populations, as we did in the
main text. We then obtain a distribution of the minimum time required to
measure the ``true" long-term growth rate (Fig.
\ref{fig:min_time_growth_dist}). We see a bimodal distribution with many
populations required 30+ years to estimate the long-term growth rate.
The large number of short years required is due to cases where the
entire time series is consistently increasing or decreasing at the same
rate.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-11-1.pdf}
\caption{Histogram of the minimum time required in order to estimate the
long-term growth rate within 20\%
error.\label{fig:min_time_growth_dist}}
\end{figure}

@@

\section{Using Generalized additive model to identify significant
trends}\label{using-generalized-additive-model-to-identify-significant-trends}

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{(a) Time series for Bigeye tuna (\emph{Thunnus obesus}) with
corresponding fitted GAM model in red and (b) statistical power as a
function of the number of years sampled. The horizontal ine at 0.8
indicates the minimum threshold for statistical power and the vertical
line denotes the minimum time required to achieve 0.8 statistical
power.\label{fig:gam_example}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-13-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance according to a GAM
model given statistical power of 0.8. The smoothing parameter was set to
3 for each population.\label{fig:min_time_dist_gam}}
\end{figure}

We then fit GAM models for 851 populations. We found a similar
distribution of minimum time required as in the main text for linear
regression (Fig. \ref{fig:min_time_dist_gam}). However, in line with our
hypothesis, the GAM models did result in a lower mean minimum time
required of 14.65 years compared to the results from the main text of
16.45 years.

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-Myhrvold2015}{}
Myhrvold, Nathan P., Elita Baldridge, Benjamin Chan, Dhileep Sivam,
Daniel L. Freeman, and S.K. Morgan Ernest. 2015. ``An amniote
life-history database to perform comparative analyses with birds,
mammals, and reptiles.'' \emph{Ecology} 96 (11): 3109.

\hypertarget{ref-GPDD2010}{}
NERC Centre for Population Biology, Imperial College. 2010. ``The Global
Population Dynamics Database Version 2.''

\hypertarget{ref-White2014}{}
White, Easton R, John D Nagy, and Samuel H Gruber. 2014. ``Modeling the
population dynamics of lemon sharks.'' \emph{Biology Direct} 9 (23):
1--18.
